Experiences with Planning for Natural Language Generation

### 2008-06-22 trunk/Papers/experiences/domains.tex
changed to Planning in NLG section (sentence generation subsection)

However, determining referring expressions (REs) and realization
interact, so it would be beneficial to perform both steps together.
This was the goal of the SPUD system \cite{Stone2003a}, which
performed both steps together using a top-down generation algorithm
based on tree-adjoining grammars \cite{joshi;etal1997} whose lexical
entries were equipped with semantic and pragmatic information.
Unfortunately, SPUD suffered from having to explore a huge search
space, and had to resort to a non-optimal greedy search strategy to
retain reasonable efficiency.

\cite{KolSto07} attempt to improve the efficiency by translating the
SPUD sentence generation problem into a planning problem and using a
planning algorithm for generation.  We will now illustrate how they do
this by a (simplified) example.

@@

Consider a knowledge base containing the individuals $r_1$ and $r_2$;
$r_1$ and $r_2$ are rabbits, $r_1$ is white and $r_2$ is brown, and
$r_1$ sleeps.  Let's say we want to express the information
$\{\mathsf{sleep}(r_1)\}$ using the tree-adjoining grammar shown in
Fig.~\ref{fig:white-rabbit-sleeps-grammar}.  This grammar consists of
\emph{elementary trees}, each of which contributes certain
\emph{semantic content}.  We can instantiate these trees by
substituting individuals for \emph{semantic roles}, such as
$\mathsf{self}$ and $\mathsf{subj}$, and then combine the tree
instances as shown in Fig.~\ref{fig:white-rabbit-sleeps-deriv} to
obtain the sentence ``The white rabbit sleeps''.  The SPUD algorithm
computes this grammatical derivation by starting with the elementary
tree for ``sleeps''; this satisfies the need to convey the semantic
information, but introduces a need to generate a noun phrase (NP) for
the subject, which also refers uniquely to the target referent $r_1$.
In a second step, it substitutes the tree for ``the rabbit'' into the
open NP leaf, which makes the derivation grammatically complete; but
as there are two different individuals that could be described as
``the rabbit'' -- technically, $r_2$ is still a \emph{distractor} for
the referring expression --, we are still not finished.  This is why
we add the tree for ``white'' to the derivation by the
\emph{adjunction} operation.  This makes the derivation syntactically
and semantically complete, and we are done.

\begin{figure}
  \centering
  xxxx
  \caption{Generating ``The white rabbit sleeps'' as a planning problem.}
  \label{fig:white-rabbit-as-planning}
\end{figure}

@@

(:action add-white
   :parameters (?u - node
                ?xself - individual)
   :precondition (and (canadjoin N ?u)
      (referent ?u ?xself)
      (rabbit ?xself))
   :effect (forall (?y - individual)
      (when (not (white ?y))
         (not (distractor ?u ?y)))))
\end{verbatim}}%
\caption{PDDL actions for generating the sentence ``The white rabbit
sleeps''.}
\label{fig:pddl-crisp}
\end{figure}

This process has clear parallels to planning: We manipulate a state by
applying actions in order to achieve a goal.  We can make this
connection precise by translating the SPUD problem into a planning
problem; the relevant excerpt of this problem is shown in
Fig.~\ref{fig:white-rabbit-as-planning}.  Each action in this planning
problem corresponds to the addition of a single elementary tree to the
derivation; the first parameter of the action is a node name in the
derivation tree, and the further parameters stand for the individuals
to which the semantic roles will be instantiated.  The syntactic
preconditions and effects are encoded using $\mathsf{open}$ and
$\mathsf{canadjoin}$ literals; the status of each referring expression
is tracked using $\mathsf{distractor}$ literals.  Notice that the
action effects contain terms of the form $\mathsf{subj}(u)$, which
construct new node names.  In order to represent the planning problem
in PDDL, these terms can be eliminated by estimating an upper bound
$n$ for the plan length, making $n$ copies of each action, ensuring
that copy $i$ can only be applied in step $i$, and replacing the term
$\mathsf{subj}(u)$ in an action copy by the constant
$\mathsf{subj}_i$.

In the example, the following plan solves the planning problem:
\begin{enumerate}
\item $\mathsf{sleeps}(\mathsf{root}, r_1)$
\item $\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}), r_1)$
\item $\mathsf{white}(\mathsf{subj}(\mathsf{root}), r_1)$
\end{enumerate}

The grammatical derivation in
Fig.~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the generated
sentence, can be systematically reconstructed from this plan.
Therefore we can solve the sentence generation problem via the detour
through planning and bring current search heuristics for planning to
bear on generation.

>>>>

In practice, the process of determining referring expressions typically
interacts with the realization step and so it turns out to be beneficial to
perform both of these steps together. This was the goal of the SPUD system
\cite{Stone2003a}, which performed both steps together using a top-down
generation algorithm based on tree-adjoining grammars \cite{joshi;etal1997}
whose lexical entries were equipped with semantic and pragmatic
information. Unfortunately, SPUD suffered from having to explore a huge
search space, and had to resort to a non-optimal greedy search strategy to
retain reasonable efficiency.

%However, determining referring expressions and realization
%interact, so it would be beneficial to perform both steps together.
%This was the goal of the SPUD system \cite{Stone2003a}, which
%performed both steps together using a top-down generation algorithm
%based on tree-adjoining grammars \cite{joshi;etal1997} whose lexical
%entries were equipped with semantic and pragmatic information.
%Unfortunately, SPUD suffered from having to explore a huge search
%space, and had to resort to a non-optimal greedy search strategy to
%retain reasonable efficiency.

@@

%\citeauthor{KolSto07}~\shortcite{KolSto07} attempt to improve the
%efficiency by translating the SPUD sentence generation problem into a
%planning problem and using a planning algorithm for generation.  We will
%now illustrate how they do this by a (simplified) example.

To improve on the efficiency of SPUD,
\citeauthor{KolSto07}~\shortcite{KolSto07} translate the sentence
generation problem into a planning problem and use a planning algorithm for
generation. We illustrate this process on a simplified example.

Consider a knowledge base containing the individuals $r_1$ and $r_2$, and a
set of attributes encoding the fact that $r_1$ and $r_2$ are rabbits, $r_1$
is white and $r_2$ is brown, and $r_1$ sleeps.  Say that we want to express
the information $\{\mathsf{sleep}(r_1)\}$ using the tree-adjoining grammar
shown in Fig.~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists
of \emph{elementary trees} (i.e., the disjoint trees in the figure), each
of which contributes certain \emph{semantic content}. We can instantiate
these trees by substituting individuals for \emph{semantic roles}, such as
$\mathsf{self}$ and $\mathsf{subj}$, and then combine the tree instances as
shown in Fig.~\ref{fig:white-rabbit-sleeps-deriv} to obtain the sentence
``The white rabbit sleeps''. The SPUD algorithm computes this grammatical
derivation by starting with the elementary tree for ``sleeps''. This tree
satisfies the need to convey the semantic information, but introduces a
need to generate a noun phrase (NP) for the subject, which also refers
uniquely to the target referent $r_1$. In a second step, SPUD substitutes
the tree for ``the rabbit'' into the open NP leaf, which makes the
derivation grammatically complete. Since there are two different
individuals that could be described as ``the rabbit''---technically, $r_2$
is still a \emph{distractor} (i.e., a \todo{one line definition here}) for
the referring expression---we are still not finished. To complete the
derivation, the tree for ``white'' is added to the existing structure by
the \emph{adjunction} operation, making the derivation syntactically and
semantically complete.

@@

(:action add-white
   :parameters (?u - node
                ?xself - individual)
   :precondition (and (canadjoin N ?u)
      (referent ?u ?xself)
      (rabbit ?xself))
   :effect (forall (?y - individual)
      (when (not (white ?y))
         (not (distractor ?u ?y)))))
\end{verbatim}}%
\caption{PDDL actions for generating the sentence ``The white rabbit
sleeps.''}
%\caption{Generating ``The white rabbit sleeps'' as a planning problem.}
\label{fig:white-rabbit-as-planning}
\end{figure}

The process described above has clear parallels to planning: we manipulate
a state by applying actions in order to achieve a goal. We can make this
connection even more precise by translating the SPUD problem into a
planning problem. For instance, Fig.~\ref{fig:white-rabbit-as-planning}
shows the corresponding PDDL actions for the above generation task, where
each action corresponds to an operation that adds a single elementary tree
to the derivation.  In each case, the first parameter of the action is a
node name in the derivation tree, and the remaining parameters stand for
the individuals to which the semantic roles will be instantiated.  The
syntactic preconditions and effects are encoded using $\mathsf{open}$ and
$\mathsf{canadjoin}$ literals; the status of each referring expression is
tracked using $\mathsf{distractor}$ literals. Notice that the action
effects contain terms of the form $\mathsf{subj}(u)$, which construct new
node names. In order to meet the syntactic requirements of PDDL, these
terms can be eliminated by estimating an upper bound $n$ for the plan
length, making $n$ copies of each action, ensuring that copy $i$ can only
be applied in step $i$, and replacing the term $\mathsf{subj}(u)$ in an
action copy by the constant $\mathsf{subj}_i$.

Once we've defined appropriate actions, we can solve the generation
problem as an ordinary planning problem. For instance, the following plan
solves our previous example:
%
\begin{enumerate}
\item $\mathsf{sleeps}(\mathsf{root}, r_1)$
\item $\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}), r_1)$
\item $\mathsf{white}(\mathsf{subj}(\mathsf{root}), r_1)$
\end{enumerate}
%
Using this plan, the grammatical derivation in
Fig.~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the generated
sentence, can be systematically reconstructed. Thus, we can solve the
sentence generation problem via the detour through planning and bring
current search heuristics for planning to bear on generation.
